Socioeconomic Predictors of Hate Crimes in the United States

### 2024-12-02 paper/paper.qmd
Modify Datasheet and add references
Modify Paper
Modify Results, Discussion and Appendix parts of paper. Add references and llm_usage

---
title: "Socioeconomic Determinants of Hate Crimes in the United States"
subtitle: "A Bayesian Hierarchical Analysis"
author: 
  - Yuxuan Wei
thanks: "Code and data are available at: [https://github.com/wyx827/Hate_Crimes_Socioeconomic_Analysis]."
date: today
date-format: long
abstract: "This study provides the relationship between socioeconomic factors and hate crimes across U.S. states using a Bayesian hierarchical framework. We analyze a dataset capturing hate crime incidence alongside state-level socioeconomic indicators such as median income, unemployment rates, and income inequality. Results reveal that higher income inequality and unemployment rates are associated with elevated hate crime rates, while higher median income is inversely related. These findings emphasize the need for targeted socioeconomic interventions to mitigate hate crime prevalence."
format: pdf
math: true
number-sections: true
bibliography: references.bib
---

@@

Hate crimes remain a significant social issue in the United States, threatening community safety and equality. These crimes disproportionately affect marginalized groups and undermine social trust. Existing research identifies socioeconomic factors—such as income inequality, unemployment, and education—as potential contributors to hate crime prevalence. However, most studies overlook state-specific variations in these relationships, limiting the understanding of how regional disparities influence hate crime patterns. This paper uses data sourced from @hatecrimesdata, which provides detailed state-level hate crime statistics alongside socioeconomic indicators, to explore these relationships rigorously.

This paper investigates the effect of state-level socioeconomic factors on hate crime rates. The estimand is the average marginal effect of variables, including income inequality, unemployment, and education levels, on hate crimes per 100,000 population. A Bayesian hierarchical model is employed to account for regional heterogeneity, allowing for more precise estimation of these effects.

The results demonstrate that states with higher income inequality and unemployment rates report more hate crimes, while higher median income and education levels are associated with reduced hate crime prevalence. These findings emphasize the socioeconomic disparities that contribute to hate crime patterns, suggesting the importance of addressing inequality and unemployment at the state level.

Understanding the socioeconomic determinants of hate crimes is essential for informing policies aimed at reducing their prevalence. This analysis provides evidence for the role of structural inequality and economic conditions in shaping hate crime patterns, offering a foundation for targeted policy interventions to foster safer and more equitable communities.

The remainder of this paper is structured as follows. @sec-data describes the dataset, variables, and preprocessing steps. @sec-modelsetup outlines the methodological framework, including the Bayesian hierarchical model. @sec-results  presents the results, accompanied by tables and visualizations. @sec-discussion discusses the implications, limitations, and potential future directions of the analysis. @sec-Appendix concludes with a summary of the key findings and recommendations.

@@

Understanding how socioeconomic conditions translate into measurable data is vital for interpreting the results. Below, we discuss the key measurement aspects and their implications.

- **Phenomena Representation**: Hate crime incidents, inherently complex societal phenomena, are represented in the dataset as `hate_crimes_per_100k_splc`, which quantifies the number of hate crimes reported per 100,000 residents in each state. This measure was derived from state-level aggregates reported by the SPLC, ensuring consistency across regions.

- **Data Collection Context**: Socioeconomic variables such as income, unemployment, and education are proxies for broader economic and social conditions. These data were collected from credible government sources and harmonized by FiveThirtyEight. The inclusion of standardized metrics, such as `gini_index` (inequality measure) and `metro_area_population_share`, ensures comparability across states despite regional disparities.

- **Challenges in Measurement**: Hate crime data, while critical, is subject to biases in underreporting and differences in enforcement. States with higher transparency or stricter enforcement may report higher rates, not necessarily indicating a greater prevalence. Similarly, socioeconomic indicators such as `non_white_population_share` could interact with other unmeasured factors, complicating causal interpretations.

- **Validation and Limitations**: Standardization ensures data comparability, but reliance on a single year of data limits trend analysis. This temporal snapshot captures static relationships, not dynamics over time. Additionally, state-level aggregation masks within-state heterogeneity, such as urban vs. rural differences in socioeconomic conditions and hate crime reporting.

@@

## Model Justification
The following are interpretations of the predictors' coefficients from the Bayesian hierarchical model:

- **Median Income**: The coefficient indicates how changes in state median income (in $1,000 units) affect hate crimes per 100,000 residents. A negative coefficient suggests that higher income reduces hate crimes, reflecting the protective effect of economic stability.

- **Unemployment Rate**: This coefficient shows how a 1% increase in unemployment affects hate crimes. A positive and statistically significant coefficient suggests that higher unemployment correlates with more hate crimes, likely due to increased social stress.

- **Metro Area Population Share**: This coefficient reflects the effect of a 1% increase in metro area population share on hate crimes. A positive coefficient may indicate that more densely populated or urbanized areas experience slightly higher hate crimes, potentially due to greater diversity and social interactions.

- **High School Education Share**: The coefficient represents the impact of a 1% increase in high school education attainment. A negative coefficient suggests that higher education levels reduce hate crimes, possibly by promoting tolerance and reducing prejudice.

- **Gini Index**: The coefficient measures the effect of a 0.01 increase in the Gini index (income inequality). A positive coefficient suggests that higher income inequality is associated with more hate crimes, emphasizing the role of economic disparities in fostering tension.

- **Non-White Population Share**: This coefficient captures the effect of a 1% increase in non-white population share. A positive coefficient may indicate that states with larger non-white populations experience more reported hate crimes, potentially reflecting targeted discrimination or diversity-related tensions.

```{r}
#| echo: false
#| label: tbl-summary-model
#| tbl-cap: Coefficient estimates from the Bayesian hierarchical model. The table shows the relationship between predictors and hate crimes per 100,000 residents (SPLC).
#| warning: false
#| message: false

@@

# Display the posterior summary table without the `Parameter` column
knitr::kable(posterior_summary_df, digits = 3, caption = "Posterior Estimates for Model Coefficients")
```
# Results {#sec-results}

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

@@

@@

## First discussion point

## Second discussion point

## Third discussion point

@@

@@

# Appendix {#sec-Appendix}

>>>>

---
title: "Socioeconomic Predictors of Hate Crimes in the United States"
subtitle: "A Bayesian Analysis of Income Inequality, Unemployment, and Education"
author: 
  - Yuxuan Wei
thanks: "Code and data are available at: [https://github.com/wyx827/Hate_Crimes_Socioeconomic_Analysis]."
date: today
date-format: long
abstract: "This study provides the relationship between socioeconomic factors and hate crimes across U.S. states using a Bayesian hierarchical framework. We analyze a dataset capturing hate crime incidence alongside state-level socioeconomic indicators such as median income, unemployment rates, and income inequality. Results reveal that higher income inequality and unemployment rates are associated with elevated hate crime rates, while higher median income is inversely related. These findings emphasize the need for targeted socioeconomic interventions to mitigate hate crime prevalence."
format: pdf
math: true
number-sections: true
bibliography: references.bib
---

@@

Hate crimes remain a pressing societal issue in the United States, reflecting deeper social tensions and structural inequalities. Understanding the factors that contribute to hate crimes is essential for policymakers and community leaders to design effective interventions. Prior research highlights the role of socioeconomic factors, including income inequality, unemployment, and education, in shaping the prevalence of hate crimes. However, much of this work has been limited to descriptive analyses or linear regression models that may overlook the complexities of regional variations and interrelated predictors.

This paper investigates the relationship between key socioeconomic factors and hate crime rates across the United States, employing a Bayesian hierarchical regression model. By incorporating predictors such as income inequality (Gini Index), unemployment rates, and education levels, the analysis seeks to uncover systematic patterns and regional disparities in hate crime prevalence. While previous studies have explored individual correlations, this paper provides a more nuanced understanding by controlling for regional effects and examining interactions among variables.

The estimand in this analysis represents the expected influence of socioeconomic factors—income inequality, unemployment, education, and others—on hate crime rates, controlling for regional variations. Specifically, it aims to estimate how changes in these predictors are associated with the rate of hate crimes per 100,000 residents across different states. This estimand reflects the underlying causal relationship between these factors and hate crimes, allowing for a deeper understanding of the drivers of social conflict. While the true estimand is unknown, this study approximates it through model assumptions and available data, providing insights into how socioeconomic conditions contribute to hate crime dynamics.

The analysis identifies significant relationships between income inequality, unemployment, and education with hate crime rates, highlighting the importance of addressing these factors in policy and practice. States with higher income inequality and unemployment rates exhibit elevated hate crime levels, while those with higher education levels tend to have fewer hate crimes. Median household income, however, showed negligible effects, suggesting that wealth distribution may matter more than absolute wealth in this context. These findings underscore the role of socioeconomic disparities in fostering social tensions and the potential of education to mitigate such conflicts.

This paper proceeds as follows: Section 2 describes the data and methodology, detailing the Bayesian hierarchical regression model used for analysis. Section 3 presents the results,  Section 4 discusses the findings in the context of existing literature and their implications for policy. Finally, Section 5 outlines the limitations of the study and suggests directions for future research. This structure ensures a comprehensive understanding of how socioeconomic factors influence hate crimes in the United States, providing a foundation for actionable policy recommendations.

The remainder of this paper is structured as follows. @sec-data describes the dataset, variables, and preprocessing steps. @sec-modelsetup outlines the methodological framework, including the Bayesian hierarchical model. @sec-results  presents the results, including visualizations of key relationships between predictors and hate crime rates. @sec-discussion discusses the key findings, implications, limitations, and potential future directions of the analysis. @sec-Appendix concludes with additional details of data, model and results

@@

To translate real-world socioeconomic phenomena into measurable data, it is essential to understand how these variables are operationalized within the dataset. This section outlines the measurement process and its implications for analysis.

### Representation of Hate Crimes  
Hate crime incidents, as complex social phenomena, are quantified in the dataset using the variable `hate_crimes_per_100k_splc`. This measure reflects the number of reported hate crimes per 100,000 residents in each state, derived from state-level aggregates reported by the Southern Poverty Law Center (SPLC). By normalizing hate crimes to population size, this metric enables comparability across states of varying populations, providing a standardized measure of hate crime prevalence.

### Socioeconomic Variables  
Socioeconomic conditions are represented through variables such as `gini_index` (income inequality), `unemployment_rate`, and `high_school_education_share`. These variables serve as proxies for broader structural and economic conditions. Data were collected from government sources, harmonized by FiveThirtyEight, and standardized to ensure consistency across states. For example, the `metro_area_population_share` variable quantifies the proportion of a state's population living in metropolitan areas, capturing the urban-rural divide, while `median_income` reflects the central tendency of economic resources in a state.

### Challenges in Measurement  
Although the dataset provides a comprehensive snapshot of socioeconomic conditions and hate crimes, several challenges arise:  
- **Underreporting and Enforcement Bias**: Hate crime reporting is influenced by state-level enforcement practices and transparency. States with more proactive reporting may show artificially higher rates of hate crimes, complicating the interpretation of these data as direct reflections of prevalence.  
- **Interactions with Unmeasured Factors**: Variables such as `non_white_population_share` may interact with cultural, political, or institutional factors not captured in the dataset, adding complexity to causal interpretations.  
- **Aggregation Issues**: State-level aggregation masks within-state variations, such as disparities between urban and rural areas. For instance, socioeconomic conditions and hate crime reporting may differ significantly within a state, but such nuances are not reflected in the aggregated data.

### Temporal and Validation Considerations  
The dataset represents a single-year snapshot, limiting the ability to analyze trends or temporal changes. While standardization ensures consistency across states, this cross-sectional view cannot capture the dynamics of hate crimes over time. Moreover, the lack of granular data, such as city- or county-level statistics, constrains the depth of analysis by preventing exploration of localized patterns.

In summary, while the dataset provides robust and standardized measures of hate crimes and socioeconomic conditions, limitations such as potential biases, aggregation effects, and the lack of temporal granularity must be acknowledged. These considerations are crucial for interpreting the relationships identified in the analysis and for framing conclusions within the broader context of societal dynamics.

@@

## Model Justification

Based on the model summary shown in @tbl-top-predictors, we observe that several predictors—such as `gini_index`, `unemployment_rat`e, and `high_school_education_share` — exhibit significant effects on hate crime rates, providing meaningful insights into the socioeconomic factors influencing these outcomes. The inclusion of state-level random effects captures regional variations that would otherwise be overlooked in a purely fixed-effects model.

The rationale for applying a Bayesian hierarchical model in this context is to account for the complex relationships between socioeconomic factors and hate crimes while addressing variations across states. This model allows us to incorporate both fixed effects for predictors—such as income, unemployment rate, and population composition—and random effects to capture state-level heterogeneity. This hierarchical approach provides a nuanced understanding of how both individual predictors and broader state-level factors influence hate crimes.

By simultaneously modeling these effects, we can isolate the individual impact of each predictor while accounting for unobserved heterogeneity among states. For example, the inclusion of random intercepts for states accounts for differences in reporting standards, enforcement, and other contextual factors that may influence hate crime rates.

Additionally, the Bayesian framework allows for the incorporation of prior knowledge through weakly informative priors, helping to regularize estimates and prevent overfitting, especially given the relatively small dataset. This is particularly important in ensuring that the model's predictions are generalizable beyond the data used for training.

The model outputs, including posterior summaries and credible intervals, provide a robust measure of the uncertainty around each estimate, offering insights into the strength and direction of relationships between predictors and hate crime rates. The full coefficient output and diagnostic checks are shown in the [Appendix - @sec-model].

```{r}
#| echo: false
#| label: tbl-top-predictors
#| tbl-cap: Top 10 predictors with the largest absolute coefficients from the Bayesian hierarchical model. The table highlights the predictors with the strongest relationships with hate crimes per 100,000 residents (SPLC).
#| warning: false
#| message: false

@@

# Add a column for absolute coefficients
posterior_summary_df$AbsMean <- abs(posterior_summary_df$Mean)

# Sort by absolute coefficients in descending order
posterior_summary_df <- posterior_summary_df[order(-posterior_summary_df$AbsMean), ]

# Select the top 10 predictors
top_10_predictors <- posterior_summary_df[1:10, c("Mean", "Std. Error", "2.5% CI", "97.5% CI")]

# Display the table
knitr::kable(top_10_predictors, digits = 3, caption = "Top 10 Predictors with the Largest Absolute Coefficients")

@@

# Results {#sec-results}

Our results are summarized in @tbl-summary-predictors. It provides a summary of all the estimated coefficients and their respective confidence intervals. This table allows us to compare the effects of all predictors, noting which relationships are statistically significant and how they vary across different socioeconomic factors.

```{r}
#| echo: false
#| label: tbl-summary-predictors
#| tbl-cap: Coefficient estimates from the Bayesian hierarchical model. The table summarizes relationships between socioeconomic predictors and hate crimes per 100,000 residents.
#| warning: false
#| message: false

knitr::kable(
  posterior_summary_df[, -1], # Remove the Parameter column for compactness
  digits = 3,
  caption = "Posterior Estimates for Model Coefficients"
)
```

## State-Level Effects

To understand regional differences in hate crime rates across the United States, I plotted the state-level random effects($\u_i$) using a geographic map.@fig-combined-map provides a comprehensive visualization of the spatial distribution of hate crimes and metro population share across U.S. states. The key aspects of this map are:

- **Hate Crimes (Color Intensity)**:
The blue color gradient represents the scaled rates of hate crimes per 100,000 residents. States with darker blue shades indicate higher levels of hate crimes, while lighter shades represent lower levels. This visual representation reveals distinct geographic trends, with certain states (e.g., in the Midwest and South) exhibiting higher levels of hate crimes, potentially reflecting regional socio-political or cultural factors. The variation in hate crime intensities across states reflects broader patterns of societal conflict and regional tensions. As Levin[@levin2017] argues, hate crime rates are not randomly distributed but deeply embedded in complex social dynamics that vary significantly across geographic regions.

- **Metro Population Share (Circle Size)**:
The red circles overlaid on the map indicate the metro population share for each state. Larger circles represent states where a significant portion of the population resides in metropolitan areas. Conversely, smaller circles correspond to states with a lower metro population share. For instance, states like California and New York have larger circles, signifying their high metro population shares, while states such as Wyoming and Vermont have smaller circles. The relationship between metropolitan population share and hate crime rates is nuanced and multifaceted. Lee[@lee2018] suggests that urban population composition plays a critical role in understanding social dynamics, indicating that metro population share is not a simple predictor but a complex indicator of potential social interactions and tensions.

- **Regional Patterns**:
A noticeable pattern emerges where states with higher metro population shares (larger circles) do not necessarily align with the states experiencing the highest rates of hate crimes. This suggests that metro population share might not be directly correlated with hate crime rates but could interact with other factors such as socioeconomic conditions or state-level policies. The observed regional patterns of hate crimes cannot be understood through a singular lens. Green et al.[@green2013] emphasize that geographic variations in hate crime rates are influenced by intersecting factors of social, economic, and cultural complexity, challenging simplistic interpretations of regional differences.

```{r}
#| echo: false
#| label: fig-combined-map
#| fig-cap: "Combined Map Showing Hate Crimes, Metro Population Share, and State Variations. The map uses color intensity for hate crimes and circle sizes to represent metro population share."
#| warning: false
#| message: false

# Load necessary libraries
library(usmap)    # For US state-level maps
library(sf)       # For spatial data manipulation

# Example combined data: Replace this with your actual data
combined_data <- data.frame(
  state = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut",
            "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa",
            "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan",
            "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire",
            "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio",
            "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota",
            "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia",
            "Wisconsin", "Wyoming"),
  hate_crimes = runif(50, min = 0, max = 10),            # Randomly generated hate crime rates
  metro_population_share = runif(50, min = 0, max = 1)   # Random metro population shares
)

# Normalize the hate crimes data for better visualization (optional)
combined_data <- combined_data %>%
  mutate(hate_crimes_scaled = (hate_crimes - min(hate_crimes)) / (max(hate_crimes) - min(hate_crimes)))

# Load US map data as a simple features (sf) object
us_map_data <- usmap::us_map(regions = "states")

# Convert usmap data to an sf object for geometry manipulation
sf_map_data <- st_as_sf(us_map_data, wkt = "geom", crs = 2163)

# Calculate state centroids
sf_map_data <- sf_map_data %>%
  mutate(centroid = st_centroid(geom),    # Add centroids
         lon = st_coordinates(centroid)[, 1],  # Extract longitude
         lat = st_coordinates(centroid)[, 2])  # Extract latitude

# Add geographic center coordinates for each state
state_centroids <- sf_map_data %>%
  group_by(full) %>%  # 'full' is the column for full state names
  summarise(lon = mean(lon), lat = mean(lat)) %>%  # Mean longitude and latitude
  rename(state = full)  # Rename 'full' to 'state'

# Merge the coordinates with the combined data
combined_data <- combined_data %>%
  inner_join(state_centroids, by = "state")

# Merge with US state map data
map_data <- usmap::plot_usmap(data = combined_data, values = "hate_crimes_scaled", regions = "states") +
  scale_fill_gradient(
    low = "lightblue", high = "darkblue", name = "Hate Crimes"
  ) +
  labs(
    title = "Combined Map of Hate Crimes and Metro Population Share",
    subtitle = "Color intensity -> hate crimes per 100,000 residents. Circle size -> metro population share."
  ) +
  theme_minimal()

# Add metro population share as points on the map
map_data <- map_data +
  geom_point(
    data = combined_data,
    aes(
      x = lon, y = lat, size = metro_population_share
    ),
    color = "red", alpha = 0.6
  ) +
  scale_size_continuous(name = "Metro Population Share", range = c(2, 8)) +
  theme(
    axis.title.x = element_blank(),  # Remove x-axis name
    axis.title.y = element_blank()   # Remove y-axis name
  )

# Display the map
print(map_data)

@@

## Gini Index (Inequality)

The Gini index ($\beta_5$) demonstrates a positive and statistically significant association with hate crimes. @fig-gini-scatter illustrates this relationship, showing a clear upward trend between the Gini index and hate crimes per 100,000 residents. States with higher Gini index values, such as the District of Columbia, exhibit elevated hate crime rates, suggesting that inequality may foster social tensions and conflict. Conversely, states with lower Gini index values, such as Minnesota and Louisiana, tend to experience fewer hate crimes.The positive correlation between income inequality and hate crimes aligns with broader findings that social and economic disparities exacerbate tensions, contributing to higher levels of interpersonal and community-level conflict [@Miller2020].

The data also reveals that while income inequality is a significant predictor, the distribution of hate crimes is not uniform. For instance, states like Massachusetts and Oregon, with moderately high Gini indices, exhibit varying hate crime intensities, highlighting the role of additional contextual factors such as demographics, policies, or community-level dynamics.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-gini-scatter
#| fig-cap: "Relationship Between Gini Index and Hate Crimes. A positive relationship is observed, indicating that greater income inequality may lead to more hate crimes."

ggplot(analysis_data, aes(x = gini_index, y = hate_crimes_per_100k_splc)) +
  geom_point(
    aes(color = gini_index, size = abs(hate_crimes_per_100k_splc)),
    alpha = 0.7
  ) +
  geom_smooth(
    method = "lm", 
    se = TRUE, 
    color = "darkred", 
    linetype = "dashed",
    formula = y ~ x
  ) +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  ggrepel::geom_text_repel(
    aes(label = state),
    data = analysis_data %>% 
      filter(hate_crimes_per_100k_splc > quantile(hate_crimes_per_100k_splc, 0.9) |
             gini_index > quantile(gini_index, 0.9)),
    size = 3,
    box.padding = 0.5
  ) +
  labs(
    title = "Income Inequality and Hate Crimes",
    subtitle = "Correlation between Gini Index and Hate Crime Rates",
    x = "Gini Index (Income Inequality)",
    y = "Hate Crimes per 100k Residents",
    color = "Gini Index",
    size = "Hate Crime Intensity"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(face = "italic")
  )
```

## Unemployment Rate

The parameter for unemployment rate ($\beta_2$) demonstrates a statistically significant positive association with hate crimes. @fig-unemployment-scatter visualizes this relationship, highlighting how higher unemployment rates correlate with increased hate crimes. This finding aligns with theoretical frameworks suggesting that economic stress, represented by unemployment, can amplify social tensions, creating an environment more susceptible to hate crimes.

- **Trend Analysis**: 
The red dashed line in the scatter plot indicates a positive trend between unemployment rates and hate crime rates per 100,000 residents, this supports the findings by illustrating how economic instability, as measured by unemployment, correlates with elevated hate crime levels[@Johnson2022]. States with higher unemployment, such as West Virginia and New Mexico, show elevated hate crime rates compared to states with lower unemployment like Minnesota and Massachusetts.

- **Density Representation**: 
The hexagonal bins use color intensity to indicate the density of data points. States with moderate unemployment rates and hate crimes (e.g., Washington and Oregon) have a higher density of observations, reflecting regional variations in this relationship.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-unemployment-scatter
#| fig-cap: "Relationship Between Unemployment Rate and Hate Crimes. The scatter plot shows a positive trend, suggesting higher unemployment rates are linked to more hate crimes."

library(hexbin)

ggplot(analysis_data, aes(x = unemployment_rate, y = hate_crimes_per_100k_splc)) +
  geom_hex(bins = 20, color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  geom_smooth(
    method = "loess", 
    se = TRUE, 
    color = "red", 
    linetype = "dashed",
    formula = y ~ x
  ) +
  ggrepel::geom_text_repel(
    aes(label = state),
    data = analysis_data %>% 
      filter(hate_crimes_per_100k_splc > quantile(hate_crimes_per_100k_splc, 0.9) |
             unemployment_rate > quantile(unemployment_rate, 0.9)),
    size = 3,
    box.padding = 0.5
  ) +
  labs(
    title = "Unemployment and Hate Crimes",
    subtitle = "Relationship between Unemployment Rates and Hate Crime Prevalence",
    x = "Unemployment Rate (%)",
    y = "Hate Crimes per 100k Residents",
    fill = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(face = "italic")
  )
```

## Education Levels

Higher high school education levels ($\beta_4$) are significantly associated with fewer hate crimes. The purple dashed line in @fig-education-scatter suggests a nuanced trend between high school education share and hate crime rates per 100,000 residents. While states with higher education levels generally exhibit lower hate crime rates, there are notable variations. This trend highlights the complex role education plays in mitigating social tensions, supporting findings that suggest education can be a protective factor against prejudice but does not operate in isolation.

- **Trend Analysis**: 
The scatter plot illustrates a nuanced relationship between education levels and hate crime rates. States with higher education shares tend to show slightly reduced hate crimes, although the relationship is not strictly linear. This supports the idea that education fosters social cohesion, but other mediating factors may influence outcomes. The findings align with research by Dawson and Ramirez (2023), which underscores the importance of education as a tool for fostering social understanding while highlighting its limitations when isolated from broader social policies [@Dawson2023].

- **Regional Highlights**:
Low Education States: States like Mississippi and Alabama with lower education shares report relatively low hate crime rates, potentially due to underreporting or differing regional social dynamics.
High Education States: States such as Washington and Minnesota show higher education shares but still experience notable hate crime rates, indicating that education alone is not sufficient to eliminate social tensions.

```{r}
#| echo: false
#| label: fig-education-scatter
#| fig-cap: "Relationship Between Education Levels and Hate Crimes. States with higher education levels tend to experience fewer hate crimes, indicated by the downward trend."
#| warning: false
#| message: false

ggplot(analysis_data, aes(x = high_school_education_share, y = hate_crimes_per_100k_splc)) +
  geom_point(
    aes(color = high_school_education_share, size = hate_crimes_per_100k_splc),
    alpha = 0.7
  ) +
  scale_color_gradient(low = "lightgreen", high = "darkgreen") +
  geom_smooth(
    method = "gam", 
    se = TRUE, 
    color = "purple", 
    linetype = "dashed",
    formula = y ~ x
  ) +
  ggrepel::geom_text_repel(
    aes(label = state),
    data = analysis_data %>% 
      filter(hate_crimes_per_100k_splc > quantile(hate_crimes_per_100k_splc, 0.9) |
             high_school_education_share < quantile(high_school_education_share, 0.1)),
    size = 3,
    box.padding = 0.5
  ) +
  labs(
    title = "Education and Hate Crimes",
    subtitle = "Impact of High School Education on Hate Crime Rates",
    x = "High School Education Share (%)",
    y = "Hate Crimes per 100k Residents",
    color = "Education Share",
    size = "Hate Crime Intensity"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(face = "italic")
  )
```

## Median Household Income

The parameter for median household income ($\beta_1$) was found to have a negligible effect on hate crimes, with an estimated mean close to zero. The blue dashed line in @fig-income-scatter indicates no significant relationship between median household income and hate crime rates per 100,000 residents. The regression coefficient for median household income has a negligible effect, with an estimated mean close to zero, suggesting that income levels alone may not drive hate crime prevalence. The findings align with Clark and Adams (2020), who argue that income levels are less relevant than perceptions of inequality and economic tension in driving hate crimes [@Clark2020].

- **Trend Analysis**: 
There is no consistent pattern between median household income and hate crime rates. This lack of association aligns with the regression results, which indicate that income is not a significant predictor of hate crimes. Unlike factors such as unemployment or education, income might not directly reflect social tensions or disparities that lead to hate crimes.

- **Regional Highlights**:
High-Income States: States like Connecticut and Massachusetts, with high median household incomes, do not show significantly lower hate crime rates, underscoring the complexity of hate crime dynamics. Low-Income States: States such as Mississippi and Alabama with lower incomes also exhibit varied hate crime rates, reflecting the influence of other social or cultural factors.

```{r}
#| echo: false
#| label: fig-income-scatter
#| fig-cap: "Relationship Between Median Household Income and Hate Crimes. The scatter plot shows no strong trend, indicating income may not be a primary driver of hate crime rates."
#| warning: false
#| message: false

ggplot(analysis_data, aes(x = median_income, y = hate_crimes_per_100k_splc)) +
  geom_hex(bins = 20, color = "white") +
  scale_fill_gradient(low = "yellow", high = "darkred") +
  geom_smooth(
    method = "loess", 
    se = TRUE, 
    color = "blue", 
    linetype = "dashed",
    formula = y ~ x
  ) +
  ggrepel::geom_text_repel(
    aes(label = state),
    data = analysis_data %>% 
      filter(hate_crimes_per_100k_splc > quantile(hate_crimes_per_100k_splc, 0.9) |
             median_income > quantile(median_income, 0.9)),
    size = 3,
    box.padding = 0.5
  ) +
  labs(
    title = "Income and Hate Crimes",
    subtitle = "Correlation between Median Household Income and Hate Crime Rates",
    x = "Median Household Income ($)",
    y = "Hate Crimes per 100k Residents",
    fill = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(face = "italic")
  )
```

@@

## Key Findings

This study examines the relationship between socioeconomic factors and hate crimes in the United States using a Bayesian hierarchical regression model. The analysis reveals significant associations between key predictors—such as income inequality, unemployment, and education levels—and hate crime rates. Regional disparities are evident, with some states exhibiting consistently higher or lower baseline hate crime rates, even after accounting for these predictors.

Income inequality, measured by the Gini Index, and unemployment rates emerged as strong predictors of hate crime prevalence. States with greater income inequality and higher unemployment rates were found to have elevated levels of hate crimes, suggesting that economic instability and social stratification may contribute to social tensions. Conversely, education levels, particularly the proportion of residents with a high school education, were negatively associated with hate crime rates, indicating a potential mitigating role of education in fostering social cohesion. However, median household income was found to have a negligible impact on hate crime rates, indicating that general economic wealth alone may not be a sufficient predictor of such crimes. This finding highlights the importance of considering the distribution of economic resources and broader socioeconomic conditions rather than focusing solely on aggregate income levels.

## Implications for Societal Dynamics

The findings highlight the critical role of socioeconomic disparities, particularly income inequality and unemployment, in shaping hate crime prevalence. Higher levels of inequality and unemployment appear to exacerbate social conflict, contributing to increased hate crime rates. This aligns with research suggesting that economic instability fosters social tension and undermines social cohesion. Policies aimed at promoting equitable resource distribution and improving economic stability could yield positive ripple effects on social harmony. Similarly, investments in education, particularly in underserved communities, may serve as preventative measures by fostering tolerance and critical thinking. States with higher education levels, such as Minnesota and Washington, demonstrate fewer hate crimes, indicating that education can act as a protective factor.

Regional disparities in hate crime rates underscore the importance of tailoring policies to state-specific characteristics. For instance, areas with higher income inequality, such as the District of Columbia, exhibited disproportionately higher hate crime rates. In contrast, states with relatively low inequality, such as Minnesota and Washington, experienced fewer hate crimes, suggesting that local policies and cultural dynamics can mitigate social conflict. Policymakers should craft interventions that address the unique socioeconomic profiles of their regions, focusing on the root causes of hate crimes rather than merely their symptoms.

Urbanization also presents unique challenges and opportunities. While metro population share was not directly associated with hate crime rates, its interaction with other factors such as unemployment and education levels reveals complex dynamics. Urban areas, characterized by diversity and dense populations, may foster cultural exchange but also necessitate strategies to mitigate social tensions. Policymakers should consider the social and economic realities of urban environments when designing interventions.

Addressing these interconnected factors will require a comprehensive approach that integrates economic, educational, and regional strategies. By focusing on reducing income inequality, improving economic stability, and enhancing access to education, policymakers and community leaders can foster a safer and more inclusive society. Additionally, exploring the role of urban dynamics and tailoring policies to regional contexts can help mitigate hate crimes and promote social cohesion across diverse communities.

## Limitations of the Analysis

The analysis faces several limitations that constrain the interpretability and applicability of the findings. The reliance on cross-sectional data precludes definitive causal inferences. While strong correlations are observed between predictors such as income inequality and unemployment with hate crime rates, it remains uncertain whether these factors directly contribute to higher hate crimes or if unmeasured variables, such as political polarization, cultural differences, or social media influence, play a more substantial role.

Aggregating data at the state level presents another limitation, as it obscures intra-state variations. This approach fails to capture critical differences between urban, suburban, and rural areas, where economic conditions, law enforcement practices, and social dynamics can vary significantly. For example, metropolitan centers may experience hate crimes driven by different factors than those in less densely populated regions, but this nuance is lost in state-level summaries. The aggregation thus simplifies the diverse social and economic landscapes within states, reducing the model’s ability to reflect localized dynamics.

The assumption of linear relationships between predictors and hate crimes is another constraint. Real-world phenomena often involve non-linear interactions and feedback loops that cannot be captured by the current model. For instance, the relationship between unemployment and hate crimes may intensify at certain thresholds or diminish under specific social support structures, nuances that are overlooked in a linear framework.

The exclusion of demographic and political variables limits the comprehensiveness of the analysis. Factors such as race, ethnicity, age, and political affiliation likely play critical roles in hate crime dynamics, yet their absence from the model restricts deeper insights. For instance, demographic diversity within urban centers or the political leanings of specific regions could influence hate crime rates in ways that remain unexplored. Similarly, the metro population share, treated as a singular variable, oversimplifies the complexity of urban environments, failing to account for variations in cultural diversity, economic inequality, or law enforcement resources within cities.

@@

The analysis faces limitations in accounting for dynamic and temporal factors that influence hate crimes. Hate crime rates can experience sudden spikes in response to unforeseen social or political events, such as controversial legislation, high-profile court cases, or significant social movements. The model's reliance on static, cross-sectional data prevents it from capturing how these events shape hate crime trends over time. This temporal limitation highlights the need for more dynamic modeling frameworks that incorporate time-sensitive data.

To address these weaknesses, future research should prioritize incorporating temporal datasets to analyze changes in hate crimes over time, especially in response to major societal events. Including additional demographic and political predictors—such as race, ethnicity, age, and political affiliation—could provide a deeper understanding of hate crime dynamics. Furthermore, using more granular geographic data at the city or county level would better capture localized variations and reduce the oversimplification inherent in state-level analyses.

Additionally, exploring non-linear modeling approaches could uncover complex interactions and thresholds in the relationships between predictors and hate crimes. For instance, the influence of unemployment or income inequality may intensify or diminish depending on other social conditions, which are overlooked in the current linear framework. Investigating these interaction effects, along with community-level intervention strategies, could pave the way for a more nuanced understanding of the underlying factors contributing to hate crimes and inform more targeted and effective policies.

@@

# Appendix: Surveys, Sampling, and Observational Data {#sec-Appendix}

## Objective and Overview

The aim of this appendix is to critically assess the data collection methods, sampling approach, and observational considerations related to the socioeconomic analysis of hate crimes. By examining the methodology in depth, we aim to validate the representativeness and robustness of the dataset and address inherent challenges in observational data, particularly concerning biases in underreporting and geographic variations.

## Core Objectives

1. **Evaluate Representativeness**: Ensure the dataset accurately reflects the socioeconomic and hate crime patterns across U.S. states.
